---
title: "Reading Data into R"
output: html_document
---

## Understanding our directory struture

First of all, let's see what is the _working directory_ that R is using at the 
moment:

```{r}
# Run this line of code by pressing Ctrl + Enter
getwd()
```

With R Markdown files, the working directory is set to the location of the file 
itself.

Our data file is located in a different folder from this document. The file 
structure is, schematically:

```
slcu_r_course
  |
  |_course_data
  |   |
  |   |_dataset_tidy.xls
  |
  |_module01_data_and_files (our working directory)
      |
      |_reading_data.Rmd
```

Therefore, to read our `CSV` file, we will need to tell _R_ to go one 
folder "above" where we are and then into the "course_data" folder. 

We can move to "one directory above" using `../`. For example:

```{r}
# This command lists the files in the directory above the current one
list.files("../")
```

**Changing the code above, can you list the files in the "course_data" folder?**

```{r}
# Complete this code
list.files("<here>")
```


## Reading data into R

To read a `CSV` file into R, we will use a function called `read_csv()`. 

This function is part of a package called `tidyverse`, which we need to load 
first:

```{r, message=FALSE}
library(tidyverse)
```

Now, try to complete the code below to read our data into R:

```{r, eval = FALSE}
# Complete this code:
surveys <- read_csv("<here>")
```

_Note:_ there is another function called `read.csv()`, which is the usual 
default function to read CSV files, and is very similar to `read_csv()`. 
During the workshops we will use `read_csv()` as it is more convenient.


