---
title: "Module 03: Data manipulation with R"
output: html_document
---

In this module we are going to see how to manipulate, subset and modify tables using `dplyr`, which is also a part of the `tidyverse` package family.

We will learn how to:

- select columns in a table
- filter a table based on certain conditions
- combine data manipulation and plotting
- summarise data across groups

First let's load the `tidyverse` package and read the data that we will work on:

```{r, message = FALSE}
library(tidyverse)

# Read data using read_csv
expt1 <- read_csv("../course_data/burghardt_et_al_2015_expt1.csv")
```

### Selecting columns with `select()`

To get an idea of data contents we will first check number of dimentsions with ``dim()`` and look at the column names with ``names()``.

```{r}
dim(expt1)
names(expt1)
```

This is quite a large dataset, we don't want to look at it all at once. Say, we are interested in just 2 columns for now: `genotype` and `temperature`. To select the columns we will use ``select()``, ironically.

```{r}
expt1.genot.temp <- select(expt1, genotype, temperature)
expt1.genot.temp
```

This was easy. Let's say we wanted to select the column `genotype` and any column that contains the word "bolt" in its name. `select()` allows you to do this:

```{r}
expt1.bolt <- select(expt1, genotype, contains("bolt"))
expt1.bolt
```

> **Challenge**: can you select all the columns that contain values measured in mm?

### Filter rows according to certain conditions with `filter()`

Let's say that we wanted to retain only the rows of our table with data on vernalized plants. This is a job for the `filter()` function.
Before the filetering we need to have a look at the possible values in the `vernalization` column, there are only 2 possibilities: 'V' and 'NV'.

```{r}
unique(expt1$vernalization)
```
Since we are intereseted in veranlized plants only, so will be selected only 
"V"s:

```{r}
expt1.vernalised <- filter(expt1, vernalization == "V")
expt1.vernalised
```

To set the filtering conditions, several operators can be used:

- `<` less than 
- `<=` less than or equal to
- `>` more than
- `>=` more than or equal to
- `==` equal to
- `!=` not equal to

It is also possible to combine several conditions together using the following operators:

- `&` for AND
- `|` for OR

We can also identify missing values (`NA`) using:

- `is.na()` is missing
- `!is.na()` is _not_ missing

For example, if we want to select individuals that were vernalized and grown 
under fluctuating temperature conditions, we could do:

```{r}
expt1.vernalised.fluctuating <- filter(expt1, vernalization == "V" & fluctuation == "Var")
```

Another example, if we wanted to select individuals grown with 8h day light or that were late flowering:

```{r}
expt1.SD.or.late <- filter(expt1, day.length == "8" | days.to.bolt > 85)
```

> **Exercise**
> Can you filter the data to keep the following?

> 1. Samples that are **not in the the Ler background** and that have been 
> treated with **fluctuating temperature**.
> 2. Samples that are bolting at less than **57 days** of flowing with **less than > 40 leaves**
> 3. Samples of genotypes `fca-6` for which `blade.ratio` is not missing


```{r}
# Enter your code here

```

### Modify or add new columns with `mutate()`

This is a useful function to add new variables (i.e. new columns) to the table, or to modify existing ones. 
Say, we want to add a new column, `blade.length.cm` which will contain blade length measeured in cm. We can construct this column by modifying the already
existing ``blade.length.mm`` column.

```{r}
dim(expt1)
expt1.cm <- mutate(expt1, blade.length.cm = blade.length.mm/10)
dim(expt1.cm)
```

We can also modify/create more than one variable at a time, separating them by `,`:

```{r}
expt1.cm <- mutate(expt1, blade.length.cm = blade.length.mm/10,
                   total.leaf.length.cm = total.leaf.length.mm/10)
```

<span style="color:red">**Be careful!**</span> If we use the name of an existing variable, it will modify it instead.

### Chain commands with `%>%` pipes 

It allows to perform a sequence of multiple operations, without a need for 
intermediate files (or very complicated nested command lines).

Now let’s say that we wanted to do several manipulations in our data.frame:

- Create a new column containing total leaf length in cm
- Filter our data so that it only contains vernalised plants
- Retain only two columns from our table: genotype and total leaf length in cm

Based on what we’ve learned so far, we could for example do the following:

```{r}
expt1.filtered <- mutate(expt1, 
                         total.leaf.length.cm = total.leaf.length.mm/10)

expt1.filtered <- filter(expt1.filtered,
                         vernalization == "V")

expt1.filtered <- select (expt1.filtered,
                          genotype,
                          total.leaf.length.cm )
```

`dplyr` allows us to “chain” several commands using a special “pipe” function `%>%`. 
This is how the above code could have been written using this trick:

```{r}
expt1 %>% 
  mutate(total.leaf.length.cm = total.leaf.length.mm/10) %>% 
  filter(vernalization == "V") %>% 
  select(genotype, total.leaf.length.cm)
```

> **Exercise:**
> Use `%>%` to do the following on expt1:

> - Keep samples that are not in the the Ler background and that have been treated with fluctuating temperature.
> - Transform the blade length in cm
> - Keep only the columns containing information about the genotype, blade length (in cm) and bolting information (hint, use `contains`)

```{r}
# Enter your code here

```


### Combine all the above with graphs

We can pipe data manipulation with `ggplot` to produce graphs for a subset of the data.
Let's say we want to make a boxplot of the flowering time in function if the 
temperature fluctuation (like in last session), however only for the vernalised samples. For this, we can use the following code: 

```{r}
expt1 %>% 
  filter(vernalization == "V") %>% 
  ggplot(aes(fluctuation, days.to.flower)) + 
  geom_boxplot()
```

> **Exercise**
> Can you make a violin plot of the days to flower for the different genotypes that have been growing in short days? 

```{r}
# Enter your code here

```

### Summarise data with `group_by` and `summarise`

Sometimes we might want to summarise our data in a smaller table. 
This kind of operation can be achieved by combining two `dplyr` functions: 
`group_by()` and `summarise()`.

For example, we can summarise the mean flowering time for each genotype:

```{r}
expt1 %>% 
  group_by(genotype) %>% 
  summarise(mean.days.to.flower = mean(days.to.flower, na.rm = TRUE))
```


### Saving data to a file

Once we have modified our data we might want to save it to come back to it later. For this we can use the function `write_csv()`

Let’s say that we were to remove all the missing data from our data to share it with a collaborator (or later use in R ourselves):

```{r}
expt1.blade.ratio.clean <- expt1 %>% 
  filter(!is.na(blade.ratio))

write_csv(expt1.blade.ratio.clean, path = "x_blade_ratio_clean.csv")
```

### Some key points to remember

- `dplyr` has several useful functions to manipulate tables
- `dplyr` functions can be “chained” together using the `%>%` pipe
- We can save tables to text-based files using the `write_*` family of functions

### Futher reading and additional resources:

- [dplyr cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [data transformation chapter from 'R for Data Science'](http://r4ds.had.co.nz/transform.html)


