---
output:
  html_document:
    toc: yes
    toc_float: yes
---

# [SLCU R Course](https://tavareshugo.github.io/slcu_r_course/)

## Data manipulation with R - material

In this module we are going to see how to manipulate, subset and modify tables 
using `dplyr`, which is also a part of the `tidyverse` package family.

We will learn how to:

- select columns in a table
- filter a table based on certain conditions
- combine data manipulation and plotting
- summarise data across groups

First let's load the `tidyverse` package and read the data that we will work on:

```{r, message = FALSE}
library(tidyverse)

# Read data using read_csv
expt1 <- read_csv("../data/burghardt_et_al_2015_expt1.csv")
```

### Selecting columns with `select()`

To get an idea of data contents we will first check number of dimentsions with 
``dim()`` and look at the column names with ``names()``.

```{r}
dim(expt1)
names(expt1)
```

This is quite a large dataset, we don't want to look at it all at once. 
Say, we are interested in just 2 columns for now: `genotype` and `temperature`. 

To select the columns we will use ``select()``:

```{r}
select(expt1, genotype, temperature)
```

Let's say we wanted to select the column `genotype` and any column that contains 
the word "bolt" in its name. `select()` allows you to do this:

```{r}
select(expt1, genotype, contains("bolt"))
```

> **Challenge**: can you select all the columns that contain values measured in mm?


### Filter rows according to certain conditions with `filter()`

Let's say that we wanted to retain only the rows of our table with data on 
vernalized plants. This is a job for the `filter()` function.

Before the filtering we need to have a look at the possible values in the 
`vernalization` column, there are only 2 possibilities: 'V' and 'NV'.

```{r}
unique(expt1$vernalization)
```

(**note:** to select a single column from a table we can use the `$` notation 
like above)

Since we are intereseted in vernalized plants only, we want to filter only those 
rows with the value "V":

```{r}
filter(expt1, vernalization == "V")
```

To set the filtering conditions, several operators can be used:

Operator  | Meaning               | Example
--------- | --------------------  | ------
`<`       | less than             | `filter(expt1, days.to.bolt < 20)`
`<=`      | less than or equal to | `filter(expt1, days.to.bolt <= 20)`
`>`       | more than             | `filter(expt1, days.to.bolt > 20)`
`>=`      | more than or equal to | `filter(expt1, days.to.bolt >= 20)`
`==`      | equal to              | `filter(expt1, days.to.bolt == 20)`
`!=`      | not equal to          | `filter(expt1, days.to.bolt != 20)`
`%in%`    | is contained in       | `filter(expt1, genotype %in% c("Col FRI", "Ler-1"))`


It is also possible to combine several conditions together using the following operators:

Operator  | Meaning               | Example
--------- | --------------------  | ------
`&`       | AND                   | `filter(expt1, days.to.bolt == 20 & genotype == "Ler-1")`
`|`       | OR                    | `filter(expt1, rosette.leaf.num < 8 | rosette.leaf.num > 100)`


We can also identify missing values (`NA`) using the `is.na()` function and its 
negation (using `!`):

Condition        | Meaning               | Example
---------------- | --------------------  | ------
`is.na()`        | is missing            | `filter(expt1, is.na(rosette.leaf.num))`
`!is.na()`       | is _not_ missing      | `filter(expt1, !is.na(rosette.leaf.num))`

For example, if we want to select individuals that were vernalized and grown 
under fluctuating temperature conditions, we could do:

```{r, eval = FALSE}
filter(expt1, vernalization == "V" & fluctuation == "Var")
```

Another example, if we wanted to select individuals grown with 8h day light or 
that were late flowering:

```{r, eval = FALSE}
filter(expt1, day.length == "8" | days.to.bolt > 85)
```

> **Challenge:**
> Can you filter the data to keep the following?
>
> 1. Samples that are **not in the the Ler background** and that have been 
> treated with **fluctuating temperature**.
> 2. Samples that bolted in less than **57 days** and with **less than 40 leaves**
> 3. Samples of genotype `fca-6` for which `blade.ratio` is not missing


### Modify or add new columns with `mutate()`

This is a useful function to add new variables (i.e. new columns) to the table, 
or to modify existing ones. 

Say, we want to add a new column, `blade.length.cm` which will contain blade length 
in _cm_ units. We can construct this column by using the already
existing ``blade.length.mm`` column:

```{r}
# Create a new table with an extra column in relation to the previous table
expt1.cm <- mutate(expt1, blade.length.cm = blade.length.mm/10)

# Check that there's now a new column
colnames(expt1.cm)
```

We can also modify/create more than one variable at a time, separating them by `,`:

```{r}
# Make both blade.length and total.leaf.length in cm:
expt1.cm <- mutate(expt1, 
                   blade.length.cm = blade.length.mm/10,
                   total.leaf.length.cm = total.leaf.length.mm/10)

# Check that those columns were now created
names(expt1.cm)
```

**Be careful!** If we use the name of an existing variable, it will modify it instead.


### Chain commands with `%>%` pipes 

"Pipes" allows us to perform a sequence of operations, without the need for 
intermediate files (or very complicated nested commands).

Let’s say that we wanted to do several manipulations in our table:

- Create a new column containing total leaf length in cm
- Filter our data so that it only contains vernalised plants
- Retain only two columns from our table: genotype and total leaf length in cm

Based on what we’ve learned so far, we could for example do the following:

```{r}
# Create new column with mutate
expt1.filtered <- mutate(expt1, 
                         total.leaf.length.cm = total.leaf.length.mm/10)

# Filter the table to retain only vernalised plants
expt1.filtered <- filter(expt1.filtered,
                         vernalization == "V")

# Select only the columns of interest
expt1.filtered <- select (expt1.filtered,
                          genotype,
                          total.leaf.length.cm )
```

`dplyr` allows us to “chain” several commands using a special “pipe” function `%>%`. 
This is how the above code could have been written using this trick:

```{r}
expt1 %>% 
  mutate(total.leaf.length.cm = total.leaf.length.mm/10) %>% 
  filter(vernalization == "V") %>% 
  select(genotype, total.leaf.length.cm)
```

What the `%>%` pipe does is take the output of the thing on the left-side of it 
and pass it on to the function that comes after it.

> **Challenge:**
> Using `%>%` to do the following on `expt1`:
>
> - Keep samples that are _not_ in the the Ler background and that have been 
treated with fluctuating temperature.
> - Transform the blade length in cm
> - Keep only the columns containing information about the genotype, blade 
length (in cm) and bolting information (hint: use `contains()`)



### Combine all the above with graphs

We can pipe data manipulation to `ggplot` to produce graphs for a subset of 
the data. This is very useful for interactive data exploration (it avoids having 
to constantly create new variables with `<-`).

Let's say we want to make a boxplot of the flowering time in function of the 
temperature fluctuation (like in last session), however only for the vernalised 
samples. For this, we can use the following code: 

```{r}
expt1 %>% 
  filter(vernalization == "V") %>% 
  ggplot(aes(fluctuation, days.to.flower)) + 
  geom_boxplot()
```

> **Challenge:**
> Can you make a violin plot of the days to flower for the different genotypes 
that have been growing in short days (8h light)? 


### Summarise data with `group_by` and `summarise`

Sometimes we might want to summarise our data into a smaller table.

This kind of operation can be achieved by combining two `dplyr` functions: 
`group_by()` and `summarise()`. Please note, grouping with `group_by()` doesn't 
change how the data looks (apart from listing how it's grouped). We can then do
summaries based on those groupings with `summarise()`.

For example, we can summarise the mean flowering time for each genotype:

```{r}
expt1 %>% 
  group_by(genotype) %>% 
  summarise(mean.days.to.flower = mean(days.to.flower, na.rm = TRUE))
```

You can see that the output contains two columns: `genotype`, which was the 
column used to "group" the original table and `mean.days.to.flower`, which was 
the column we created with the `summarise` function. And there's only 10 rows 
of this new table, which corresponds to one row for each genotype.

You can group the table by more than one variable, and you can also make more 
than one summary operation on the table. For example, we might want to calculate 
both the mean, median and standard deviation of the days to flower of each genotype 
grown in long and short days:

```{r}
expt1 %>% 
  group_by(genotype, day.length) %>% 
  summarise(mean.days.flower = mean(days.to.flower, na.rm = TRUE),
            sd.days.flower = sd(days.to.flower, na.rm = TRUE),
            median.days.flower = median(days.to.flower, na.rm = TRUE))
```

Now, you can see there's 20 rows in that table. This is because we have one row 
for each combination of genotype (there's 10 of them) and day.length (8h and 16h).

It is often useful to find how many observations we have for each group. 

In the example above, let's say we wanted to know how many plants we have of 
each genotype. For this we can use the special function `n()`, which will count 
the number of rows of each group in the original table:

```{r}
expt1 %>% 
  group_by(genotype) %>% 
  summarise(mean.days.to.flower = mean(days.to.flower, na.rm = TRUE),
            n.obs = n())
```
Please note, you can use ``n()`` only inside ``summarise()`` function. There is
a similar function, ``count()`` that can be used independently and piped in.
Because counting the number of observations in a group is such a common operation,
there's actually a special function, `count()` just to do that one thing:

```{r}
expt1 %>% 
  count(genotype, day.length)
```

Often, we want to apply the same set of summary functions to several variables. 
For example, in our dataset we might want to group the data by genotype and treatment 
groups and then calculate the mean and standard deviation for all the traits 
measured. This kind of operation can be achieved using the `summarise_at` function.

Here is how it works:

```{r}
# Summarise all traits by treatment groups and genotype
expt1 %>% 
  group_by(genotype, background, temperature, fluctuation, day.length, vernalization) %>% 
  summarise_at(vars(days.to.bolt, days.to.flower, rosette.leaf.num, cauline.leaf.num, blade.length.mm, total.leaf.length.mm, blade.ratio), 
               funs(mean, sd)) %>% 
  ungroup()
```

**Note:** It is important to say that, after you use the `group_by()` function, 
the table remains grouped by the variables you defined. This means that any operations 
you do afterwards will use this grouping, which might not be desirable! There is a 
function called `ungroup()`, which is usually good to use after you finish your 
operations.

### Saving data to a file

Once we have modified our data we might want to save it to come back to it later. 
For this we can use the function `write_csv()`

Let’s say that we were to remove some missing data from our data to share 
it with a collaborator (or later use in R ourselves):

```{r, eval = FALSE}
expt1.blade.ratio.clean <- expt1 %>% 
  filter(!is.na(blade.ratio))

#alternatively, we can use drop_na() function that we used in the module02:
expt1.blade.ratio.clean2 <- expt1 %>% drop_na(blade.ratio)

write_csv(expt1.blade.ratio.clean, path = "../data/expt1_blade_ratio_clean.csv")
```


### Some key points to remember

- `dplyr` has many many useful functions to manipulate tables
- `dplyr` functions can be “chained” together using the `%>%` pipe
- We can save tables to text-based files using the `write_*` family of functions


### Futher reading and additional resources:

- [dplyr cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [data transformation chapter from 'R for Data Science'](http://r4ds.had.co.nz/transform.html)


