---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# [SLCU R Course](https://tavareshugo.github.io/slcu_r_course/)

## Reading data into R - solution to exercise

>### Exercise:
> 
> Throughout the course we will be using Rstudio, which is a program that provides 
with a convenient interface to work with R. 
>
> R code can be saved into text documents, called scripts (usually with a file 
extension `.R`). Writing code into scripts is ideal, because you keep a record 
of all the operations that you did during your data analysis. This code can then 
be re-run, modified, and shared for reproducible data analysis!
>
>- Open the file "reading_data.Rmd" in the "module01_data_and_files" course folder
>- Try and complete the code following the instructions in that document

The full solution is given here, and an explanation follows below:

``` r
# First load all the libraries needed for this script
library(tidyverse)

# Check what the working directory is at the moment
#  this is what "folder" R is working from at the moment
getwd()

# Set working directory to the course materials folder
# Change this to be suitable to your situation
setwd("/home/slcu.user/Desktop/slcu_r_course/module01_data_and_files")

# Read the CSV file and store the table in an object called "my_surveys"
## In this case, the CSV file is located in the "data" folder
my_surveys <- read_csv("data/dataset_tidy.csv")

# View the loaded table
View(my_surveys)
```

### Exercise explanation

When you start an analysis script, it's always best to load all the necessary 
packages right at the top of your code - this way anyone using this code will 
know what packages they need installed in order for the code to work:

```{r, message=FALSE}
library(tidyverse)
```

Then, it's good to see what is the _working directory_ that R is using 
at the moment

```{r, eval=FALSE}
getwd()
```

The output of that command will vary depending on your operating system and 
username. 

For example, for a user called "slcu.user", this might be the default working directory 
on a Mac:

`/Users/slcu.user/`

And this one on a Windows:

`C:/Users/slcu.user/Documents`

When doing an analysis, it's best to set the working directory to be in the 
folder where your project is stored in. 

Let's say that the course materials are on your Desktop. So, schematically, this 
is where your data file is located:

```
Desktop
  |
  |_slcu_r_course
      |
      |_module01_data_and_files
          |
          |_data
             |
             |_dataset_tidy.csv
```

Therefore, it's best if you first change your working directory to be in 
`module01_data_and_files`. 

For our previous example this might be something like:

```{r, eval=FALSE}
# On a Mac
setwd("/Users/slcu.user/Desktop/slcu_r_course/module01_data_and_files")

# On a Windows
setwd("C:/Users/slcu.user/Desktop/slcu_r_course/module01_data_and_files")
```

--------

Now that you've set the working directory, you can read the data in by using 
the function called `read_csv()`:

```{r, eval=FALSE}
surveys <- read_csv("data/dataset_tidy.csv")
```

_Note:_ there is another function called `read.csv()`, which is the usual 
default function to read CSV files, and is very similar to `read_csv()`. 
During the workshops we will use `read_csv()` as it is more convenient.


